Testing traversability of protein tunnels with flexible ligands using motion planning

### 2017-06-14 paper.tex
Updates to tunnel properties

\def\L{L}
\def\LA{L_1}
\def\LB{L_2}

@@

To guide thee growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of a sphere, around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to ``skip'' such parts of the tunnels that are
not traversible or reachable by the ligand.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}main loop of the RRT planner}
\KwIn{
    tunnel $T=(t_i, r_i), n=1,\ldots,N$ represented by $N$ spheres,
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 0$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < N$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

@@

    \For{$i= N-1,N-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $t_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal
scale.
First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.

@@

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories in order to compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
  \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
\end{itemize}
Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
Moreover, we map the representant's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

@@

The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}).
The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
The tunnel is represented by 760 spheres with average distance $0.05$~\AA.

% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).

\begin{figure}
\centering
\caption{\label{fig::tunnel}
    TODO image of the tunnel + 1cqw
}
\end{figure}

\begin{table}
\centering
\caption{\label{tab::rrt}
    Runtime and sucess ratio of the planner
}
\input table.003.tex
\hskip 10pt
\input table.004.tex
\end{table}

Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.

>>>>

\def\L{L}

@@

To guide thee growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of a sphere, around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to ``skip'' such parts of the tunnels that are
not traversible or reachable by the ligand.
The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}main loop of the RRT planner}
\KwIn{
    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 0$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < N$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

@@

    \For{$i= N,N-1,\ldots v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree to $T[i]$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with maximal
scale.
For a given conformation $l \in \L$, first the maximal scale $\smax$ is considered and
$m$ samples are generated around $\qrand$ and tested for collision.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.

@@

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
  \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
\end{itemize}
During visualization, we map these properties onto a tunnel's surface so that they can be visually correlated with, e.g., the tunnel's shape.

Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
Moreover, we map the representant's cluster size onto the width of the polyline.
In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).

@@
